Fusing Filters with Integer Linear Programming

### 2014-05-14 2014betterfusionforfilters/section/04-ILP.tex
varied fixes

\begin{figure}[H]
\begin{tabbing}
MMMM        \= MM   \= \kill
$nodes$     \> @::@ \> $program \to V$          \\
$edges$     \> @::@ \> $program \to E$          \\
$edge$      \> @::@ \> $\{bind\} \times bind \to E$\\
\\
MMMMM       \= \kill
$nodes(bs)$ \> $= \{(name(b), \tau_b) | b \in bs\}$       \\
\end{tabbing}
\end{figure}
Each node may simply be the name of its output binding (or bindings, in the case of @external@); as we require names to only be bound once, this is assured to be unique.
Creating edges between these nodes is simply when one binding references an earlier one. The only complication is designating edges as \emph{fusible} or \emph{fusion-preventing}.

\begin{figure}[H]
\begin{tabbing}
MMMMM       \= \kill
$edges(bs)$ \> $= \bigcup_{b \in bs}edge(b)$    \\
\\
MMMM             \= M \= \kill
$edge(bs, out = @fold@~f~in)$ \\
                              \> $=$    \> $\{inedge(bs,out,s) | s \in f\}$ \\
                              \> $\cup$ \> $\{inedge(bs, out, in) \}$       \\
$edge(bs, out = @map@~f~in)$  \\
                              \> $=$    \> $\{inedge(bs,out,s) | s \in f\}$ \\
                              \> $\cup$ \> $\{inedge(bs, out, in) \}$       \\
$edge(bs, out = @filter@~f~in)$  \\
                              \> $=$    \> $\{inedge(bs,out,s) | s \in f\}$ \\
                              \> $\cup$ \> $\{inedge(bs, out, in) \}$       \\
$edge(bs, out = @gather@~data~indices)$  \\
                              \> $=$    \> $\{(out,data, fusion~preventing) \}$ \\
                              \> $\cup$ \> $\{inedge(bs, out, indices) \}$       \\
$edge(bs, out = @cross@~a~b)$            \\
                              \> $=$    \> $\{inedge(bs, out, a) \}$       \\
                              \> $\cup$ \> $\{(out, b, fusion~preventing) \}$ \\
$edge(bs, outs = @external@~ins)$  \\
                              \> $=$    \> $\{(outs,i, fusion~preventing) | i \in ins \}$ \\
\\
$inedge(bs,to,from)$ \\
                     \> $|$ \> $(from = @fold@~f~s) \in bs$     \\
                     \> $=$ \> $(to, from, fusion~preventing)$  \\
                     \> $|$ \> $(\ldots,from,\ldots = @external@ \ldots) \in bs$     \\
                     \> $=$ \> $(to, from, fusion~preventing)$  \\
                     \> $|$ \> $otherwise$                      \\
                     \> $=$ \> $(to, from, fusible)$            \\
\end{tabbing}
\end{figure}

@@

\begin{tabbing}
M   \= MM \= MMMMMMM \= MM \= \kill
$x$   \> @:@  \> $node \times node$ \> $\to$ \> $\mathbb{B}$
\end{tabbing}
The first and most important group is created for every pair of nodes, and has a boolean value indicating whether the two nodes can be fused. If $x_{ij} = 0$, then $i$ and $j$ are fused together. While it may seem counterintuitive for 0 to mean fused and 1 unfused, it means the objective function can simply minimise a weighted $x_{ij}$, and also makes later precedence constraints simpler. The values of these variables are used to construct the partitioning at the end, such that $\forall i,j.\ x_{ij} = 0 \iff cluster(i) = cluster(j)$. \ben{When we're talking about the dependency graph, when we say ``nodes can be fused'', it means that the operators that produce these values can be fused -- or that the values are produced by the same cluster of operators.}

@@

\begin{tabbing}
M   \= MM \= MMMMMMM \= MM \= \kill
$c$   \> @:@  \> $node$             \> $\to$ \> $\mathbb{B}$
\end{tabbing}
The third, and final, group of variables are purely for minimisation, indicating whether a node's array is \emph{fully contracted}.
They are not required for correctness of the clustering, but are used in the objective function.
A node's array will be fully contracted --- that is, it will be replaced by a scalar --- if all nodes that use this array are fused together: $c_i = 0 \iff \forall (i,j) \in E. x_{ij} = 0$. \ben{Using $\forall$ like this implies that the quantified $i$ isn't the same as the one attached to $c_i$. Write this with a list comprehension instead.}

% -----------------------------------------------------------------------------
\subsubsection{Unoptimised version}
Before showing the optimised version with certain constraints removed (\S\ref{s:OptimisedConstraints}), this simpler, unoptimised version is shown. The only difference is that fewer constraints and variables are required in the optimised version, but both versions are equivalent. \ben{What does ``equivalent'' mean when the constraints are different?} \ben{Say what ``optimised'' means in this case.} \ben{Give some hints about why the unoptimised version has more constraints. }

@@

\begin{figure}[H]
\begin{tabbing}
MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
Minimise   \> \ldots \\
Subject to \> \ldots \\
           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
           \> (if $\tau(i) \not= \tau(j) \wedge parents(i,j) = \emptyset$)  \\
           \\
           \> $x_{i'j}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> $x_{ij'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> $x_{i'j'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> (if $\tau(i) \not= \tau(j) \wedge parents(i,j) = \{i',j'\}$) \\
           \> \ldots
\end{tabbing}
\end{figure}
If two nodes have different iteration sizes, they may not necessarily be fused.
However, if they have parent transducers of the same size, they may be fused if both their parents are fused together, and each is fused with its respective parent.

@@

\paragraph{Array contraction}
\begin{figure}[H]
\begin{tabbing}
MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
Minimise   \> \ldots \\
Subject to \> \ldots \\
           \> $x_{ij}$    \> $\le$ \> $c_i$           \>       \>            \\
           \> (for all edges from i)            \\
           \> \ldots \\
\end{tabbing}
\end{figure}
As per Darte's work on optimising for array contraction\cite{darte2002contraction}, we define a variable $c_i$ for each array.
An array is fully contracted if all its output edges are fused with it. Thus, $c_i=0$ only if $\forall j | (i,j) \in E. x_{ij} = 0$. By minimising $c_i$ in the objective function, we favour solutions that reduce the number of required intermediate arrays. \ben{Say what array contraction is, and why we want this constraint.}

@@

\subsection{A note on transitivity}
It may seem that we could generate far fewer constraints and rely on transitivity of clustering equality $x_{ij}$.
This means that for each pair $i,j$ we wouldn't generate an $x_{ij}$ and thus not generate those constraints, unless $i$ or $j$ is a @filter@, or arc between $i$ and $j$.
This would help, but also means we can't \emph{minimise} on these removed $x_{ij}$, so the solution won't count the (rather smaller) benefits of fusing two non-connected nodes.

\subsubsection{Fusion-preventing path optimisation}
\label{s:OptimisedConstraints}
Alternatively:
we can say, only generate constraints if there is no \emph{fusion preventing path} between $i$ and $j$.
Here, we generate fewer constraints than originally, but still more than above.
We also must preprocess the graph to find such blocking paths.
But we still get the minimisation to count non-connected nodes.

\begin{figure}[H]
\begin{tabbing}
MMMMMM      \= MM   \=  MMMMMMMMMMMM    \=  \kill
$split$     \> @::@ \> $V \times E \to \{ \{name\} \}$      \\
MMMM        \= M    \= \kill
$split(vs,es)$ \\
    \> $=$  \> $\{clusterable(v,vs,es) | v \in vs\}$      \\
$clusterable(v,vs,es)$  \\
    \>$=$       \>$\{v' | v' \in vs$                                \\
    \>$\wedge$  \>$\forall p. p = path(v,v') \vee p = path(v',v)$   \\
    \>$\wedge$  \>$fusion~preventing \not\in p\}$\\
\end{tabbing}
\end{figure}

We can simplify the version above and remove many constraints and variables by noting that if there is a fusion-preventing edge between $i$ and $j$, $x_{ij} = 1$, so the precedence constraint can be simplified to just $\pi_i < \pi_j$, and the $x_{ij}$ variable removed.
Likewise with clusters of different, incompatible types.
The contraction variable $c_i$ can be removed, if $i$ has any fusion-preventing output edges: any such edges use the array and cannot be fused, thus there is no possibility of the array being contracted away.

\begin{figure}[H]
\begin{tabbing}
MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
Minimise   \> \ldots \\
Subject to \> $-N x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
           \>             (no edge from i to j, but there is a fusion benefit)            \\
           \>    $x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
           \>             (an edge from i to j)            \\
           \>             \>       \> $\pi_i < \pi_j$ \>       \>            \\
           \>             (a \emph{fusion-preventing} edge from i to j)            \\
\\
           \> $x_{ij}$    \> $\le$ \> $c_i$           \>       \>            \\
           \> (for all edges from i with no fusion-preventing outputs)      \\
\\
           \> $x_{i'j}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> (if $\tau_i \not= \tau_j \wedge gen_i=\{i'\}$) \\
           \> $x_{ij'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> (if $\tau_i \not= \tau_j \wedge gen_j=\{j'\}$) \\
\end{tabbing}
\end{figure}

>>>>

\begin{figure}[H]
\begin{tabbing}
MMMM        \= MM   \= \kill
$nodes$     \> @::@ \> $program \to V$          \\
$edges$     \> @::@ \> $program \to E$          \\
$edge$      \> @::@ \> $\{bind\} \times bind \to E$\\
\\
MMMMM       \= \kill
$nodes(bs)$ \> $= \{(name(b), \iiter_{\Gamma,C}(b)) | b \in bs\}$       \\
\end{tabbing}
\end{figure}
Each node may simply be the name of its output binding (or bindings, in the case of @external@); as we require names to only be bound once, this is assured to be unique.
Creating edges between these nodes is simply when one binding references an earlier one. The only complication is designating edges as \emph{fusible} or \emph{fusion-preventing}.

\begin{figure}[H]
\begin{tabbing}
MMMMM       \= \kill
$edges(bs)$ \> $= \bigcup_{b \in bs}edge(bs, b)$    \\
\\
MM             \= M \= \kill
$edge(bs, out = @fold@~f~in)$ \\
    \> $=$    \> $\{inedge(bs,out,s) | s \in fv(f)\} \cup \{inedge(bs, out, in) \}$       \\
$edge(bs, out = @map@~f~in)$  \\
    \> $=$    \> $\{inedge(bs,out,s) | s \in fv(f)\} \cup \{inedge(bs, out, in) \}$       \\
$edge(bs, out = @filter@~f~in)$  \\
    \> $=$    \> $\{inedge(bs,out,s) | s \in fv(f)\} \cup \{inedge(bs, out, in) \}$       \\
$edge(bs, out = @gather@~data~indices)$  \\
    \> $=$    \> $\{(out,data, \fusionpreventing) \} \cup \{inedge(bs, out, indices) \}$       \\
$edge(bs, out = @cross@~a~b)$            \\
    \> $=$    \> $\{inedge(bs, out, a) \}           \cup      \{(out, b, \fusionpreventing) \}$ \\
$edge(bs, outs = @external@~ins)$  \\
    \> $=$    \> $\{(outs,i, \fusionpreventing) | i \in ins \}$ \\
\\
$inedge(bs,to,from)$ \\
    \> $|$ \> $(from = @fold@~f~s) \in bs$     \\
    \> $=$ \> $(to, from, \fusionpreventing)$  \\
    \> $|$ \> $(outs = @external@ \ldots) \in bs     \wedge from \in outs$     \\
    \> $=$ \> $(to, outs, \fusionpreventing)$  \\
    \> $|$ \> $otherwise$                      \\
    \> $=$ \> $(to, from, \fusible)$            \\
\end{tabbing}
\end{figure}

@@

\begin{tabbing}
M   \= MM \= MMMMMMM \= MM \= \kill
$x$   \> @:@  \> $node \times node$ \> $\to$ \> $\mathbb{B}$
\end{tabbing}
The first and most important group is created for every pair of nodes, and has a boolean value indicating whether the two nodes can be fused. If $x_{ij} = 0$, then $i$ and $j$ are fused together. While it may seem counterintuitive for 0 to mean fused and 1 unfused, it means the objective function can simply minimise a weighted $x_{ij}$, and also makes later precedence constraints simpler. The values of these variables are used to construct the partitioning at the end, such that $\forall i,j.\ x_{ij} = 0 \iff cluster(i) = cluster(j)$.
If two nodes are ``fused together'', it means that their combinators will be in the same cluster.

@@

\begin{tabbing}
M   \= MM \= MMMMMMM \= MM \= \kill
$c$   \> @:@  \> $node$             \> $\to$ \> $\mathbb{B}$
\end{tabbing}
The third, and final, group of variables are purely for minimisation, indicating whether a node's array is \emph{fully contracted}.
They are not required for correctness of the clustering, but are used in the objective function.
A node's array will be fully contracted --- that is, it will be replaced by a scalar --- if all nodes that use this array are fused together: $c_i = 0 \iff \forall (i',j) \in E.\ i = i' \implies x_{ij} = 0$.

% -----------------------------------------------------------------------------
\subsubsection{Unoptimised version}
Before showing the optimised version with certain constraints removed (\S\ref{s:OptimisedConstraints}), this simpler, unoptimised version is shown.
The only difference is that fewer constraints and variables are required in the optimised version, but both versions give the same clustering.
This unoptimised version generates clustering constraints and variables for every pair of nodes, regardless of whether they may, in fact, be fused together.
Later, we show that certain constraints and variables can be removed when there is a fusion-preventing edge between two nodes.

@@

\begin{figure}[H]
\begin{tabbing}
MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
Minimise   \> \ldots \\
Subject to \> \ldots \\
           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
           \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge parents(i,j) = \emptyset$)  \\
           \\
           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
           \> (if $\bot \in \{\iiter_{\Gamma,C}(i), \iiter_{\Gamma,C}(j)\}$)  \\
           \\
           \> $x_{i'j}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> $x_{ij'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> $x_{i'j'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> (if $\tau(i) \not= \tau(j) \wedge parents(i,j) = \{i',j'\}$) \\
           \> \ldots
\end{tabbing}
\end{figure}
If two nodes have different iteration sizes, they may not necessarily be fused.
If either of the nodes have no iteration size --- in which case, they are @external@ computations --- they may not be fused at all.
However, if they have parent transducers of the same size, they may be fused if both their parents are fused together, and each is fused with its respective parent.

@@

\paragraph{Array contraction}
\begin{figure}[H]
\begin{tabbing}
MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
Minimise   \> \ldots \\
Subject to \> \ldots \\
           \> $x_{ij}$    \> $\le$ \> $c_i$           \>       \>            \\
           \> (for all edges from i)            \\
           \> \ldots \\
\end{tabbing}
\end{figure}
As per Darte's work on optimising for array contraction\cite{darte2002contraction}, we define a variable $c_i$ for each array.
An array is fully contracted if all its output edges are fused with it.
Thus, $c_i=0$ only if $\forall (i',j) \in E.\ i = i' \implies x_{ij} = 0$.
By minimising $c_i$ in the objective function, we favour solutions that reduce the number of required intermediate arrays.

If an array is fully contracted, it means that the array does not need to be materialised in memory, and can simply be replaced by a scalar variable for the current element. Contracting arrays reduces memory traffic and can reduce overall memory usage.

@@

\subsubsection{A note on transitivity}
It may seem that we could generate far fewer constraints and rely on transitivity of clustering equality $x_{ij}$.
This means that for each pair $i,j$ we wouldn't generate an $x_{ij}$ and thus not generate those constraints, unless $i$ or $j$ is a @filter@, or arc between $i$ and $j$.
This would help, but also means we can't \emph{minimise} on these removed $x_{ij}$, so the solution won't count the (rather smaller) benefits of fusing two non-connected nodes.

\subsubsection{Fusion-preventing path optimisation}
\label{s:OptimisedConstraints}
We can remove certain constraints and variables, by only checking nodes if there is no \emph{fusion-preventing path} between $i$ and $j$.
If there is a fusion-preventing path between two nodes, they are known to be in different clusters.
We can simplify the integer linear program, and make the job of the ILP solver easier, by ommitting these constraints.

\begin{figure}[H]
\begin{tabbing}
MMMMMM      \= MM   \=  MMMMMMMMMMMM    \=  \kill
$split$     \> @::@ \> $V \times E \to \{ \{name\} \}$      \\
MMMM        \= M    \= \kill
$split(vs,es)$ \\
    \> $=$  \> $\{clusterable(v,vs,es) | v \in vs\}$      \\
$clusterable(v,vs,es)$  \\
    \>$=$       \>$\{v' | v' \in vs$                                \\
    \>$\wedge$  \>$\forall p \in path(v,v') \cup path(v',v).\ \fusionpreventing \not\in p\}$\\
\end{tabbing}
\end{figure}

\begin{figure}[H]
\begin{tabbing}
MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
Minimise   \> \ldots \\
Subject to \> $-N x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
           \>             (no edge from i to j, but there is a fusion benefit)            \\
           \>    $x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
           \>             (an edge from i to j)            \\
           \>             \>       \> $\pi_i < \pi_j$ \>       \>            \\
           \>             (a \emph{fusion-preventing} edge from i to j)            \\
\\
           \> $x_{ij}$    \> $\le$ \> $c_i$           \>       \>            \\
           \> (for all edges from i with no fusion-preventing outputs)      \\
\\
           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
           \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge parents(i,j) = \emptyset$)  \\
           \\
           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
           \> (if $\bot \in \{\iiter_{\Gamma,C}(i), \iiter_{\Gamma,C}(j)\}$)  \\
           \\
           \> $x_{i'j}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> $x_{ij'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> $x_{i'j'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> (if $\tau(i) \not= \tau(j) \wedge parents(i,j) = \{i',j'\}$) \\
\end{tabbing}
\end{figure}
